---
title: "Mixture models with *mixglm*"
subtitle: "Contents"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{to_delete}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
```
*mixglm* is an [R](https://cran.r-project.org/) package for fitting mixture 
models. This vignette documents its use.\
\

# Installation
The package *mixglm* uses NIMBLE which requires a compiler such as [Rtools.exe](https://cran.r-project.org/bin/windows/Rtools/). 
For installation instructions see: [NIMBLE](https://r-nimble.org/download).

To install the development version of the mixglm package from [GitHub](https://github.com/) use:
```{r installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("adamklimes/mixglm")
```
# Introduction
*mixglm* uses Bayesian framework via [NIMBLE](https://r-nimble.org/) to fit 
mixtures of regressions. It provides flexibility in specification of these 
regressions allowing their mean, precision, and probability to be dependent 
on up to multiple predictors. It supports [normal](https://en.wikipedia.org/wiki/Normal_distribution), [beta](https://en.wikipedia.org/wiki/Beta_distribution), [gamma](https://en.wikipedia.org/wiki/Gamma_distribution), or [negative binomial](https://en.wikipedia.org/wiki/Negative_binomial_distribution) 
distributed response variables.\
\
Mixtures of regressions can be used for various purposes. Mixture models can
be used as a semiparametric way for modelling variables with an unknown 
distribution (<a name="bMcLachlan"></a>[McLachlan & Peel, 2000](#refMcLachlan)). 
They can be used in cluster analyses, or in resilience modelling (ecology) for 
parametrization of stability landscapes.\
\
The *mixglm* package provides tools for exploration of results such as 
computations and visualizations of mixtures. While these tools are usable for
all above mentioned purposes, they are described in terminology of resilience 
modelling in this vignette.

## *mixglm* for resilience modelling
Resilience is often illustrated using a ball rolling on a curve, where the ball 
represents a system which rolls into a valley representing a *stable state*. 
Due to a disturbance or internal stochasticity, the ball can be pushed up from 
the stable state. Resilience can be then defined as either local or non-local 
stability of the ball on the curve. If the slopes of the valley are steep, the 
system is locally stable and highly resilient 
(so called *engineering resilience*). If the nearest hill (local maximum of 
the curve, a *tipping point*) is far from the system position, then 
the system is stable also non-locally and has high *ecological resilience* 
(see review <a name="bDakos"></a>[Dakos & Kéfi, 2022](#refDakos)).\
\
This *stability curve* can change its shape along external predictors creating 
a so-called *stability landscape* (see <a name="bKlimes"></a>[Klimeš et al., 2024](#refKlimes)). Mixture of 
regressions can be used to parametrize this stability landscape. From it, 
stability curves for predictor values of interest can be explored and the 
position of instances of a system (e.g. observed ecological systems) along 
these curves can be evaluated. *mixglm* offers tools for that ranging from 
stability landscape calculation and visualization, depiction of specified 
stability curves, to calculation of several metrics of ecological resilience.\
\

# Fitting mixture models using *mixglm*
In *mixglm*, it is possible to fit mixtures with a predefined number of 
components that have normal, beta, gamma, or negative binomial distribution 
(all components have the same distribution). For each component, its mean, 
precision, and probability within the mixture can be defined as being dependent 
on predictors which can differ between components (and also between mean, 
precision, probability sub-model). In the following we present a step-by-step 
guide for fitting a mixture model using *mixglm*.

## Package set-up and data preparation
We start by loading the *mixglm* package and generating data for the example. 
We generate two example variables, one representing a precipitation gradient 
and the other being a proxy for ecosystem state - such as vegetation height. To 
fully illustrate the functionality of *mixglm*, vegetation height will
have a multimodal distribution for a part of the precipitation gradient 
representing two alternative states.

```{r dataset}
library(mixglm)

st <- function(x, y = x) (x - mean(y)) / sd(y)  # standardization function

set.seed(10)
n <- 300
state <- rbinom(n, 1, 0.5)
precip <- runif(n, c(200, 600)[state + 1], c(1600, 2000)[state + 1])
vegHeight <- rnorm(n, c(1000, 2000)[state + 1], 100)
precipSt <- st(precip)
vegHeightSt <- st(vegHeight)
dat <- data.frame(vegHeight, precip, vegHeightSt, precipSt)

plot(vegHeight ~ precip, data = dat, ylab = "Vegetation height (cm)")
```

We generated an auxiliary latent variable `state` which describes to which of 
the two ecosystem states any particular observation belongs. We assumed 
vegetation height to be normally distributed with parameters depending on this 
`state`. 

## Fitting the mixture model
To fit a mixture model, we need to specify three sub-models. The first sub-model
(`stateValModels`) specifies the mean values of mixture components, the second 
(`statePrecModels`) their precision, and the third (`stateProbModels`) their 
probability within the mixture. Each of them is specified by a [formula](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/formula) with
predictors on the right side.\
\
In resilience modelling, we typically do not have a priori reasons to 
differentiate predictors between mixture components nor between sub-models. 
This means that mean, precision, and probability of each component can change 
along each predictor (see below; but one might want to limit the flexibility to
facilitate the fit, especially with modest datasets - e.g. by constraining the
precision to be constant: `statePrecModels = ~ 1`).\
\
We also need to specify the number of components in the mixture (`numStates`). 
In resilience modelling, this number should at least correspond to the number of 
hypothesized alternative stable states. However, multiple components can 
together represent one stable state, therefore it is advisable to select 
a higher number of components. The number of components can be determined using
the Watanabe-Akaike information criterion (WAIC; <a name="bWatanabe"></a>[Watanabe, 2013](#refWatanabe)) by running the 
model with different numbers and selecting the model with the lowest WAIC.\
\
In our example, we will use just two components for simplicity. We also use 
standardized response variable and predictor to mean 0 and standard deviation 1. Standardization is recommended since it facilitates model fitting (especially 
finding starting values for the sampler and specification of informative priors 
if desired). We also specify the error distribution as `gaussian` in our case.

```{r model}
mod <- mixglm(
  stateValModels = vegHeightSt ~ precipSt,
  statePrecModels = ~ precipSt,
  stateProbModels = ~ precipSt,
  stateValError = "gaussian",
  inputData = dat,
  numStates = 2,
  mcmcChains = 2,
  setSeed = TRUE
)
```

Fitting mixture models is challenging. Various mixtures often fit the same data 
similarly well and samplers can easily get stuck in local optima. To 
evaluate the fit, e.g. the [coda](https://CRAN.R-project.org/package=coda) 
package and its tools can be used - here we illustrate by plotting sampled 
values for individual parameters. On the left, we see sampled values of the 
specified parameter for each chain providing visual clue about convergence and 
on the right we see posterior probability density of the parameter:

```{r coda}
plot(mod$mcmcSamples$samples[, "intercept_statePrec[1]"])
```

*mixglm* offers a `plot()` function which shows the mean, standard deviation, 
and probability of each component along one predictor. Each chain is visualized
separately by default (`byChains = TRUE`) providing also visual clues about 
convergence. 

```{r plot}
plot(mod)
```

## Model specifications
*mixglm* allows different predictors for individual components. For each 
sub-model (`stateValModels/statePrecModels/stateProbModels`), we can provide
a list of formulas with each formula used for one component. This can be used
e.g. to explore model behavior or test specific hypotheses. 

```{r flists, eval = FALSE}
statePrecModels = list(~ precipSt, ~ 1) # precision of 2. component is constant
```

Priors can be specified using the `setPriors` argument. It takes a list of up 
to three named lists, one for each sub-model. We can specify priors for 
intercept parameters and priors for slope parameters. 
`int1` denotes the intercept of the first component and `int2` denotes intercept 
of all other components. Here we show several examples of setting priors:

```{r priors, eval = FALSE}
# Full specification of priors
setPriors = list(               
  stateVal = list(             # dnorm(mean, tau) parametrization is used
    int1 = "dnorm(0.0, 0.01)", #   by default, where sd = 1 / sqrt(tau)  
    int2 = "dgamma(1.0, 1.0)", # dgamma(shape, rate) parametrization is used
    pred = "dnorm(0.0, 0.01)"), 
  stateProb = list(
    int2 = "dnorm(0.0, sd = 10.0)",  # alternative parametrization can be used
    pred = "dnorm(0.0, 0.01)"),
  statePrec = list(
    int = "dnorm(0.0, 0.01)",
    pred = "dnorm(0.0, 0.01)"))

# Specification of priors for regression slope (effect of predictors on mean
#   value of each component). All unspecified priors are set to defaults. 
setPriors = list(               
  stateVal = list(
    pred = "dnorm(0.0, 0.1)"))

# Specification of all intercept priors. All unspecified priors are set to 
#   defaults. 
setPriors = list(               
  stateVal = list(             
    int1 = "dnorm(5.0, 1.0)",  
    int2 = "dbeta(1.0, 20.0)",
  stateProb = list(             
    int2 = "dnorm(0.0, 0.1)", 
  statePrec = list(            
    int = "dnorm(0.0, 0.1)"))
```

The `mixglm()` function has six arguments to control the MCMC run. Arguments 
`mcmcIters`, `mcmcBurnin`, `mcmcChains`, and `mcmcThin` can be used to set the 
number of iterations for each MCMC chain, the number of initial iterations to be 
discarded, the number of MCMC chains, and the thinning interval respectively. 
`setSeed` sets R's random number seed. The argument `setInit` can be used to 
specify starting values for the sampler. This can be useful for exploration of 
parameter space or to ensure valid starting values. Initial values of a model 
can be accessed by `mod$initialValues`.

## Results
Apart from the `plot()` function which visualizes individual components along a
predictor, *mixglm* offers several other functions for accessing and plotting
results.

### Accessing results
Calling the model object (`mod` in our case) prints WAIC and posterior mean 
values for all parameters. These values can be accessed using the `coef()` 
function. For a more detailed overview, `summary()` function provides also the 
standard deviation, 2.5, 25, 75, and 97.5 percentiles of the posterior 
distribution for each parameter. Using the argument `byChains = TRUE`, we get 
this summary separately for each chain.\
\
The `summary()` function has two more useful arguments. Intercepts for the mean 
values of components are in *mixglm* always modeled as a sum of all intercept 
parameters with the same or lower rank (in combination with positive priors for 
intercept parameters with rank > 1, this is to ensure identifiability of the 
model). Therefore, intercept parameters for the `stateVal` sub-model do not
represent intercepts of individual components (but their cumulative sum does).
To get intercepts of individual components, set parameter `absInt = TRUE`. 
Finally, the argument `randomSample` can be used to take random samples of 
parameters from their posterior distributions.

```{r summary}
summary(mod, absInt = TRUE)
```

### Plotting results
*mixglm* has two functions which visualize resulting mixture distributions.
The function `landscapeMixglm()` shows a heatplot presentation of the fitted 
mixture with reddish colours denoting high probability density (in resilience 
modelling called *stability landscape*). It plots the probability density 
scaled to range from 0 to 1 and highlights local maxima (suggesting stable 
states in resilience modelling; blue by default) and local minima (tipping 
points; red by default). 

```{r landscape}
landscapeMixglm(mod, axes = FALSE, xlab = "Precipitation (mm/yr)", 
  ylab = "Vegetation height (cm)")
axis(1, labels = 1:3*500, at = st(1:3*500, precip))    # non-standardized 
axis(2, labels = 1:4*500, at = st(1:4*500, vegHeight)) #   values on axes 
```

In the case there are multiple predictors, the argument `form` can be used to 
specify which predictor to visualize on the horizontal axis, and the argument
`otherPreds` to specify values of predictors which are not visualized.
Be aware that (as for ordinary multiple regression) for multiple 
predictors, observations (black points) are projected to the visualized plane 
and thus cannot be used to assess the fit of the model to the data.\
\
Often the resulting mixture has small local minima or maxima which do not 
representmajor peaks on the probability density function. If minima and maxima 
of the probability density are of interest (as in resilience modelling), it is 
usefulto set a threshold and consider only major bumps in the probability 
density function as actual minima/maxima. The argument `threshold` can be used 
to denote only such minima/maxima which differ in the scaled probability 
density by at least specified value.

```{r threshold}
landscapeMixglm(mod, threshold = 0.3)
```

The argument `randomSample` of the `landscapeMixglm()` function can be used to 
assess uncertainty of the results. It specifies how many random samples from 
the posterior distribution should be taken. For each of these samples, the 
probability density of the mixture is evaluated and the standard deviation of 
these evaluations is plotted alongside local minima and maxima for each of 
them. These computations can take some time, thus we illustrate it here using 
just 3 samples:

```{r uncertainty}
landscapeMixglm(mod, threshold = 0.3, randomSample = 3)
```

The second plotting function is `sliceMixglm()`. It can be understood as a 
vertical slice through the plot from the `landscapeMixglm()` or `plot()` 
functions. It shows the probability density for a given predictor value 
(argument `value`; in resilience modelling called stability curve). It also 
highlights the mean of each mixture component (by default using the same 
colours as the `plot()` function). As in the `plot()` function, `sliceMixglm()` 
has an argument `byChains` (which is by default `TRUE`) showing the probability 
density and mean of mixtures for each chain (thus highlighting improper 
convergence of chains).

```{r curve}
sliceMixglm(mod, value = 0.5) # precipSt == 0.5
```

The argument `addEcos` can be used to include observations with similar 
predictor values (differing up to `ecosTol`, by default 0.1).

```{r addEcos}
sliceMixglm(mod, value = 0.5, addEcos = TRUE, ecosTol = 0.3)
```

The argument `randomSample` is used instead of the posterior mean to take and 
plot random samples from the posterior distributions of parameters. 

## Predictions
To predict the scaled probability density of the mixture for specific predictor 
values or to get local minima/maxima for those values, *mixglm* has the 
function`predict()`. By providing predictor values as a named data.frame to 
the `newdata` argument, we get a list with a `probCurves` item which stores 
the scaled probability density for each predictor value along the response 
variable (response variable values are stored in item `sampledResp`). Item 
`tipStable` provides a data.frame for each provided predictor value of local 
minima (`state == 0`) and maxima (`state == 1`) of the probability density. 
Those which pass a user defined threshold (specified by the `threshold` 
argument) are marked as `catSt == 1`. The scaled probability density of these 
local minima and maxima and their response values are also provided.

```{r predict}
pred <- predict(mod, newdata = data.frame(precipSt = c(-1.5,0,1.5), 
  vegHeightSt = c(1, 1, 1)), threshold = 0.3)

# we can e.g. produce plots similar to sliceMixglm()
plot(pred$probCurves$X1 ~ pred$sampledResp, type = "l")

# we add highlighting of local minima/maxima and if they reach the threshold
tipStable1 <- pred$tipStable[[1]]
cols <- c("red", "blue")
points(tipStable1$resp, tipStable1$probDens, col = cols[tipStable1$state + 1],
 pch = c(1, 16)[tipStable1$catSt + 1])
```

## Resilience measures
The function `predict()` also provides several resilience measures. Potential 
energy (`potentEn`) is a scaled probability density for each observation. 
`distToState` is distance to the closest stable state (local maxima of 
probability density) taking `threshold` into account. Distance is in units of 
the response variable and can be understood as a vertical distance of an 
observation to a blue line in the `landscapeMixglm()` figure. For observations
with tipping points on their stability curve (local minima of probability 
density), distance to closest tipping point (`distToTip`) and potential depth
(`potentialDepth`) is provided. Potential depth is the difference in potential 
energy between an observation and the closest tipping point.

```{r resilience, fig.height = 7}
# prediction for the original (modeled) data can be done by omitting the 
#   argument `newdata`
predOrig <- predict(mod, threshold = 0.3) 
par(mfrow = c(2,2))                       
plot(dat$precip, dat$vegHeight, cex = predOrig$obsDat$potentEn * 2, 
  main = "Potential energy")
plot(dat$precip, dat$vegHeight, cex = predOrig$obsDat$distToState * 4, 
  main = "Distance to stable state")
plot(dat$precip, dat$vegHeight, cex = predOrig$obsDat$distToTip, 
  main = "Distance to tipping point")
plot(dat$precip, dat$vegHeight, cex = predOrig$obsDat$potentialDepth * 1.5, 
  main = "Potential depth")

```

# References
[↑](#bDakos) <a name="refDakos"></a>Dakos, V., Kéfi, S., 2022. Ecological resilience: what to measure and how. *Environmental Research Letters* 17: 043003. [doi.org/10.1088/1748-9326/ac5767](https://doi.org/10.1088/1748-9326/ac5767)\
\
[↑](#bKlimes) <a name="refKlimes"></a>Klimeš, A., Chipperfield, J., Töpper, Macias-Fauria, J., Spiegel, M., Vandvik, V., Velle, L. G., Seddon, A. [mixglm]: A tool for modelling ecosystem resilience. *bioRxiv*. [doi.org/10.1101/2024.03.23.586419](https://doi.org/10.1101/2024.03.23.586419)\
\
[↑](#bMcLachlan) <a name="refMcLachlan"></a>McLachlan, G., Peel, D., 2000. Finite Mixture Models. *A Wiley-Interscience Publication*. John Wiley & Sons, Inc. ISBN 0-471-00626-2\
\
[↑](#bWatanabe) <a name="refWatanabe"></a>Watanabe, S. 2013. A widely applicable bayesian information criterion. *Journal of Machine Learning Research* 14: 867–897. [Link](https://jmlr.org/papers/v14/watanabe13a.html)
