---
title: "Case Studies"
subtitle: "Contents"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
```
This is an article illustrating the use of *mixglm* package on 2 case studies.
For more technical illustration of the package functionality, see 
`vignette("mixglm")` (*mixglm* has to be installed with vignettes).\
\

# Tree cover in South America
The first case study replicates analyses by Hirota et al. 
(<a name="bHirota"></a>[2011](#refHirota)) and Flores et al. 
(<a name="bFlores"></a>[2024](#refFlores)). These two studies model tree cover 
for the most of South America and for the Amazon respectively along 
precipitation gradient. The used tree cover data is from 2001. The replication 
is done using tree cover for the year 2025 for whole South America. \
\

## Tree cover dataset
*mixglm* package provides dataset `treeCover` which includes tree cover, 
precipitation, temperature, predicted precipitation for the end of 21. century 
and coordinates for 5,000 data-points across South America. 

```{r dataset}
library(mixglm)

# see ?treeCover for additional information 
str(treeCover)
```

In this case study, we will be modelling stability landscape along precipitation
and temperature gradients. There are only few observations with very high
precipitation and very low temperature. That prevents us from estimating
reliably the stability landscape for these parts of respective gradients. 
Therefore, we will constrain the analyses to the well represented parts of 
precipitation (below 4000 mm/yr) and temperature gradients (over 0°C).

```{r dataSelection}
plot(treeCover$treeCover ~ treeCover$precip, cex = 0.1, ylab = "Tree cover (%)",
  xlab = "Precipitation (mm/yr)")
abline(v = 4000, lty = 2)
plot(treeCover$treeCover ~ treeCover$temp, cex = 0.1, ylab = "Tree cover (%)",
  xlab = "Temperature (°C)")
abline(v = 0, lty = 2)
datTC <- treeCover[treeCover$precip < 4000 & treeCover$temp > 0, ] # 4860 observations
```

## Stability landscape along precipitation gradient
First, let's model stability landscape along precipitation gradient. Tree cover 
has upper (100%) as well as lower (0%) limit, therefore we will use not normal 
but beta mixture. Beta-distributed variable can have values from 0 to 1 
(excluding 0 and 1). Thus we calculate proportion of tree cover and shift the 
values slightly to make sure the dataset does not contain any exact zeros or 
ones. We also standardize predictors to 0 mean and standard deviation 1 to 
facilitate the model fit (and ensure the same effect of priors once we use 
multiple predictors). 

```{r dataPreparation}
squeeze <- function(x) x * 0.98 + 0.01
datTC$treeCoverProp <- squeeze(datTC$treeCover / 100)

st <- function(x, y = x) (x - mean(y)) / sd(y)
datTC$precipSt <- st(datTC$precip)
datTC$tempSt <- st(datTC$temp)
```

We need to specify the number of components in the mixture. We should use more
components than is the number of expected stable states because each stable 
state can be modeled using multiple components (resulting e.g. in wider basin). 
Since probability of each component in the mixture is estimated, unnecessary 
components will have probability close to zero and will be effectively 
suppressed. We could select the number of components based on an Watanabe-Akaike 
information criterion (WAIC; <a name="bWatanabe"></a>[Watanabe, 2013](#refWatanabe))
by running the model with different numbers of components and selecting the 
one with the lowest WAIC. However, estimates of WAIC are sometimes unstable 
for mixture models. In conclusion, we recommend using a high number of 
components and ensure that results are not effected by adding or excluding a 
component. In this example, we use 7 components. We also limit the number of 
chains to just 2 (for illustration purposes) to make the fit quicker (following
model fit can take up to a few hours). 

```{r model}
numStates <- 7
set.seed(21)
mod <- mixglm(
  stateValModels = treeCoverProp ~ precipSt,
  stateProbModels = ~ precipSt,
  statePrecModels = ~ precipSt,
  stateValError = "beta",
  inputData = datTC,
  numStates = numStates,
  mcmcChains = 2
)
```

### Exploration of results
To explore the results, we start with plotting individual components. It shows us
how they fit the data and by using the default argument `byChains = TRUE`, we can
see if chains did not converge well. For example, below we can see double lines 
for some mixture components showing that the convergence was not perfect in 
this case. For further discussion of convergence and how to assess it, 
see `vignette("mixglm")`.\
\
Upper part of the figure shows probability of each component along the 
precipitation gradient. Lower part depicts mean and standard deviation for each 
component along the same gradient.

```{r plot}
plot(mod, cex = 0.2)
```

To get the overall stability landscape, we use the `landscapeMixglm` function. 
By default, it shows stable states in blue and tipping points in red with scaled
probability density as a background. Small "bumps" in the stability landscape 
unlikely denote real stable states. To avoid them, we set `threshold` value to
be higher than zero.

```{r landscape}
landscapeMixglm(mod)
# Threshold = 0.1 and original precipitation values (x-axis)
landscapeMixglm(mod, threshold = 0.1, axes = FALSE, xlab = "Precipitation (mm/yr)",
  ylab = "Tree cover (%)")
axis(2, labels = 0:5*20, at = squeeze(0:5/5), las = 2)
axis(1, labels = c(0:4*1000), at = st(c(0:4*1000), datTC$precip))
```

The same function can be used to visualize uncertainty by plotting stability 
landscape based on random posterior samples. Background shows standard deviation 
of these landscapes. Since the plotting can be slow with higher number of 
samples, we use only 5 here.

```{r uncertainty}
landscapeMixglm(mod, threshold = 0.1, randomSample = 5)
```

We can explore stability curves for any precipitation value using the 
`sliceMixglm` function. Since we used standardized precipitation as a predictor,
we need to standardize the precipitation value for which we want the stability 
curve. Apart from the stability curve, the figure shows also estimates of each 
component at the specified precipitation value.

```{r slice}
sliceMixglm(mod, value = st(2000, datTC$precip))
```

### Maps of domains and resilience
Since the stability landscape can be complex with theoretically many stable 
states, there is no general way to assign observations to domains of individual
stable states. In this case study, however, it is relatively simple since we 
identified two clearly distinct stable states. First, we use function `predict`
to get stable state and tipping points for all modeled observations 
(along with resilience measures which we will use later). Function `predict` 
calculates stability curve for each observation (modeled or provided as 
a `newdata`) and returns its stable states (`state == 1`) and tipping points 
with scaled probability density, response variable value (`treeCoverProp`), 
and whether they satisfy the threshold (`catSt == 1`). 

```{r domains}
pred <- predict(mod, threshold = 0.1)
# Stable states and tipping points for observation 3
tipStable3 <- pred$tipStable[[3]]
tipStable3

# Stability curve for observation 3 with stable states (blue) and 
#   tipping points (red) (full circles when they satisfy the threshold)
plot(pred$probCurves$X3 ~ pred$sampledResp, type = "l", xlab = "treeCoverProp",
  ylab = "Probability density (scaled)")
cols <- c("red", "blue")
points(tipStable3$resp, tipStable3$probDens, col = cols[tipStable3$state + 1],
  pch = c(1, 16)[tipStable3$catSt + 1])

# A function to determine the domain (for this case study!)
isDomainForest <- function(tipStable, resp){
  # Extract response of stable states/tipping points which satisfy the threshold
  stateResp <- tipStable$resp[tipStable$state > 0.5 & tipStable$catSt == 1]
  tipResp <- tipStable$resp[tipStable$state < 0.5 & tipStable$catSt == 1]
  stateResp <- stateResp[!is.na(stateResp)]
  tipResp <- tipResp[!is.na(tipResp)]
  # Auxiliary function to provide minimum of x or Inf if x has length 0
  minE <- function(x) if (length(x) == 0) Inf else min(x)
  # Response value of domain's stable state
  domainStateResp <- c(stateResp[stateResp <= resp & resp - stateResp < minE((resp - tipResp)[resp - tipResp > 0])],
                 stateResp[stateResp > resp & stateResp - resp < minE((tipResp - resp)[tipResp - resp > 0])])
  if (length(domainStateResp) < 1) NA else
    # Domain with response value of stable state above 0.6 is always forest
    if (domainStateResp > 0.6) 1 else 0
}
domainForest <- mapply(isDomainForest, pred$tipStable, datTC$treeCoverProp)

# Visuall check
landscapeMixglm(mod, threshold = 0.1, addPoints = FALSE)
points(datTC$precipSt, datTC$treeCoverProp, 
  col = c("lightgreen", "darkgreen")[domainForest + 1])

# Map of domains
plot(datTC$x, datTC$y, col = c("lightgreen", "darkgreen")[domainForest + 1], 
  pch = 15, xlab = "Longitude (°)", ylab = "Latitude (°)", asp = 1, cex = 0.5)
```

We can similarly plot various resilience measures stored in `pred$obsDat`:

```{r resilience}
# A function to assign colors to a continuous measure
assignColor <- function(dat, col){
  sq <- seq(min(dat, na.rm = TRUE), max(dat, na.rm = TRUE), length.out = length(col)+1)
  col[findInterval(dat, sq, all.inside = TRUE)]
}

# Map of resilience (potential depth)
potentialDepthCol <- assignColor(pred$obsDat$potentialDepth, heat.colors(130)[1:100])
plot(datTC$x, datTC$y, col = "grey", pch = 15, xlab = "Longitude (°)", 
  ylab = "Latitude (°)", asp = 1, cex = 0.5)
points(datTC$x, datTC$y, col = potentialDepthCol, pch = 15, cex = 0.5)
```

## Stability landscape with multiple predictors
Package *mixglm* allows to specify multiple predictors for each mixture 
component (and also for mean, precision, and probability). This is
desirable because relevant external conditions (such as climate) cannot be 
often captured using a single predictor. Downside is that it increases 
dimensionality of the model which then cannot be easily visualized in 2D. To 
illustrate *mixglm* model with multiple predictors, we add temperature as a 
predictor to our case study of tree cover in South America. 

```{r modelPT}
set.seed(21)
modPT <- mixglm(
  stateValModels = treeCoverProp ~ precipSt + tempSt,
  stateProbModels = ~ precipSt + tempSt,
  statePrecModels = ~ precipSt + tempSt,
  stateValError = "beta",
  inputData = datTC,
  numStates = nStates,
  mcmcChains = 2
)
```


# References
[↑](#bFlores) <a name="refFlores"></a>Flores, B.M., Montoya, E., Sakschewski, B. et al. 2024. Critical transitions in the Amazon forest system. *Nature* 626, 555–564. [doi.org/10.1038/s41586-023-06970-0](https://doi.org/10.1038/s41586-023-06970-0)\
\
[↑](#bHirota) <a name="refHirota"></a>Hirota, M., M. Holmgren, E. H. van Nes, and M. Scheffer. 2011. Global resilience of tropical forest and savanna to critical transitions. *Science* 334: 232–235. [doi.org/10.1126/science.1210657](https://doi.org/10.1126/science.1210657)\
\
[↑](#bWatanabe) <a name="refWatanabe"></a>Watanabe, S. 2013. A widely applicable bayesian information criterion. *Journal of Machine Learning Research* 14: 867–897. [Link](https://jmlr.org/papers/v14/watanabe13a.html)
